---
title: "Cell communication"
output: html_notebook
---

```{r}
library(Seurat)
library(nichenetr)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
library(DESeq2)
```

# Load the data
```{r}
memory.limit(size=90000)
seurat <- readRDS("C:/.../Full dataset.rds")
```

#Load DSA results table
```{r}
result_table <- read.csv("C:/.../dsa_results_ANOVA.csv",stringsAsFactors = F)
```


# Prepare NicheNet prior model
```{r}
ligand_target_matrix <- readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns
```
```{r}
lr_network <- readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
head(lr_network)
```
```{r}
weighted_networks <- readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
weighted_networks_lr = weighted_networks$lr_sig %>% inner_join(lr_network %>% distinct(from,to), by = c("from","to"))
```



```{r}
seurat$GA <- factor(seurat$GA)
```
```{r}
levels(seurat$GA) <- c("GA_14.1", "GA_14.3", "GA_15.1",  "GA_15.5",  "GA_16.0", "GA_16.3", "GA_17.5", "GA_18.2", "GA_19.0")
```
```{r}
seurat$celltype <- seurat$seurat_clusters
```


# Differential state analysis
```{r}
library(muscat)
```

#Add 1 factor for all clusters
```{r}
seurat$AllClusters <- "NA"
seurat$AllClusters[seurat$seurat_clusters %in% c("0", "1", "2", "3", "4", "5", "6", "7", "8")] <- "AllCells"
```


# Run NicheNet <<<-
-> Start with GA_14.1 vs GA_19.0 -> Play with logFC and frequencies.
```{r}
clusters <- result_table %>%
  filter(padj <= 0.01 &
  abs(fval) >= 0) %>%
  select(Cluster) %>%
  group_by(Cluster) %>%
  summarise(nGene = n())
clusters
```

```{r}
receiver_clusters <- filter(clusters, nGene >= 100) %>%
  pull(Cluster)
receiver_clusters
```

```{r}
getFrequency <- function(cluster){
  cluster_cells <- colnames(seurat)[seurat$seurat_clusters == cluster]
  
  subset_mat <- as.matrix(seurat[["RNA"]]@data[, cluster_cells])
  subset_frq <- rowSums(subset_mat > 0) / ncol(subset_mat)
  genes_keep <- rownames(subset_mat)[subset_frq >= 0.1]
  return(genes_keep)
}
```

```{r}
runNicheNet <- function(receiver){
  print(paste0("Performing analysis on cluster: ", receiver))
  #define receiver
  expressed_genes_receiver <- getFrequency(receiver)
  background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
  
  #define senders
  sender_celltypes <- clusters$Cluster[which(clusters$Cluster %in% result_table$Cluster)]
  list_expressed_genes_sender <- sender_celltypes %>% unique() %>% lapply(getFrequency) 
  expressed_genes_sender <- list_expressed_genes_sender %>% unlist() %>% unique()
  
  #get Hyx genes of interest
  geneset_oi <- filter(result_table, padj <= 0.01 & 
                      abs(fval) >= 0 &
                       Cluster == receiver)  %>% 
                       pull(Gene)
  #geneset_oi <- filter(geneset_oi, GA_14.1.frq >= 0.1 | GA_19.0.frq >= 0.1) %>% pull(gene) <-<-
  
  #define potential ligands for genes
  ligands <- lr_network %>% pull(from) %>% unique()
  receptors <- lr_network %>% pull(to) %>% unique()
  expressed_ligands <- intersect(ligands,expressed_genes_sender)
  expressed_receptors <- intersect(receptors,expressed_genes_receiver)
  potential_ligands <- lr_network %>% 
    filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% 
    pull(from) %>% 
    unique()
  
  #perform nichenet ligand activity analysis: rank potential ligands based on the presence
  #of their target genes in the gene set of interest (compared to the background set of genes)
  ligand_activities <- predict_ligand_activities(geneset = geneset_oi, 
                                              background_expressed_genes = background_expressed_genes, 
                                              ligand_target_matrix = ligand_target_matrix, 
                                              potential_ligands = potential_ligands)
  ligand_activities <- ligand_activities %>% arrange(-pearson) %>% mutate(rank = rank(dplyr::desc(pearson)))
  ligand_activities$receiver_cluster <- receiver
  
  return(ligand_activities)
}
```

```{r}
Sys.time()
ligand_activities <- lapply(receiver_clusters, runNicheNet) 
Sys.time()
```

```{r}
ligand_activities <- do.call("rbind", ligand_activities)
```

```{r}
saveRDS(ligand_activities, file="C:/.../ligand_activities.rds")
```

```{r}
ligand_activities <- readRDS("C:/.../ligand_activities.rds")
```

```{r}
pearson_scores <- ggplot(ligand_activities, aes(x=receiver_cluster, y=pearson)) + 
  geom_jitter(size=0.25) + 
  theme_bw() + 
  theme(axis.text.x=element_text(angle=45, hjust=1))
pearson_scores
```

```{r}
seurat$celltype <- seurat$seurat_clusters
```



# Ligand plot functions
## Ligand activity
# Summary figures for each subset
```{r}
subset_dict <- data.frame(subset = seurat$AllClusters,
                          CellType = seurat$celltype)
subset_dict <- unique(subset_dict)
subset_dict <- separate(subset_dict, subset, c("subset", "clusterID"))
subset_dict$clusterID <- NULL
subset_dict$subset <- as.character(subset_dict$subset)
subset_dict$AllClusters <- as.character(subset_dict$CellType)
```

```{r}
ligand_auroc_heatmap <- function(subset_name, plot_width, plot_height){
  
  subset_clusters <- subset_dict %>% filter(subset == subset_name) %>% pull(CellType)
  
  ligands <- ligand_activities %>%
    filter(receiver_cluster %in% subset_clusters &
           rank <= 20)
  
  ligand_mat <- ligands[,c("test_ligand", "auroc", "receiver_cluster")]
  ligand_mat <- pivot_wider(ligand_mat, names_from="receiver_cluster", values_from="auroc")
  ligand_mat <- as.data.frame(ligand_mat)
  ligand_list <- ligand_mat$test_ligand
  cell_types <- colnames(ligand_mat)[2:ncol(ligand_mat)]
  ligand_mat <- as.matrix(ligand_mat[,2:ncol(ligand_mat)])
  rownames(ligand_mat) <- ligand_list
  colnames(ligand_mat) <- cell_types
  ligand_mat[is.na(ligand_mat)] <- 0
  ligand_mat[ligand_mat > 0] <- 1
  
  #Can't do clustered columns if only one cell type
  if(length(unique(ligands$receiver_cluster)) > 1){
    ligand_heatmap <- pheatmap(ligand_mat,
         color = c("whitesmoke", "firebrick"),
         legend = F,
         cluster_rows=T,
         cluster_cols=T,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         clustering_method="ward.D2",
         border_color = "black",
         filename=paste0("C:/.../", 
                         subset_name, "_ligand_activity.png"),
         width = plot_width,
         height = plot_height)
  } else{
    ligand_heatmap <- pheatmap(ligand_mat,
         color = c("firebrick", "firebrick"),
         breaks = c(0,1),
         legend = F,
         cluster_rows=F,
         cluster_cols=F,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         clustering_method="ward.D2",
         border_color = "black",
         filename=paste0("C:/.../", 
                         subset_name, "_ligand_activity.png"),
         width = plot_width,
         height = plot_height)
  }
  
  return(ligand_heatmap)
}
```

```{r}
unique(subset_dict$subset)
```

```{r}
ligand_all <- ligand_auroc_heatmap("AllCells", 2, 12)
```


```{r}
Idents(seurat) <- "seurat_clusters"
```

## Average expression
```{r}
getAvgExp <- function(cluster, gene_order){
  cells <- colnames(seurat)[seurat$seurat_clusters == cluster & seurat$GA == "GA_19.0"]
  avg <- rowMeans(as.matrix(seurat[["RNA"]]@data[gene_order,cells]))
  return(avg)
}
```

```{r}
ligand_expression_heatmap <- function(ligand_heatmap, subset_name,
                                      plot_height){
  
  #Get ligand list from heatmap
  gene_order <- ligand_heatmap$gtable$grobs[[3]]$label
  
  clusters <- levels(Idents(seurat))
  ligand_avg <- lapply(clusters, getAvgExp, gene_order=gene_order)
  ligand_avg <- do.call("cbind", ligand_avg)
  colnames(ligand_avg) <- clusters
  
  ligand_exp_heatmap <- pheatmap(ligand_avg,
         color = colorRampPalette(c("whitesmoke", "purple"))(100),
         breaks = seq(0, 2, length.out=101),
         cluster_rows=F,
         cluster_cols=F,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         clustering_method="ward.D2",
         border_color = "black",
         filename=paste0("C:/..../", subset_name, "_ligand_expression.png"),
         width=7, 
         height=plot_height)
  
}
```

```{r}
ligand_expression_heatmap(ligand_all, "AllCells", 10)
```

## Fold change
```{r}
getFoldChange <- function(cluster, gene_order){
  filtered_table <- filter(result_table, Cluster == cluster &
                             padj <= 0.01)
  
  fc <- filtered_table$fval
  names(fc) <- as.character(filtered_table$Gene)
  fc <- fc[gene_order]
  names(fc) <- gene_order
  fc[is.na(fc)] <- 0
  
  return(fc)
}
```

```{r}
ligand_fc_heatmap <- function(ligand_heatmap, subset_name, plot_height){
  gene_order <- ligand_heatmap$gtable$grobs[[3]]$label
  clusters <- levels(Idents(seurat))
  
  fold_change <- lapply(clusters, getFoldChange, gene_order = gene_order)
  
  fold_change <- do.call("cbind", fold_change)
  colnames(fold_change) <- clusters
  
  ligand_fc_heatmap <- pheatmap(fold_change,
         color = colorRampPalette(brewer.pal(8, "Blues"))(100),
         breaks = seq(-0, 20, length.out=101),
         cluster_rows=F,
         cluster_cols=F,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         clustering_method="ward.D2",
         border_color = "black",
         filename=paste0("C:/..../", subset_name, "_ligand_foldchange.png"),
         width=7, height=plot_height)
}
```

```{r}
ligand_fc_heatmap(ligand_all, "AllCells", 10)
```

# Putative receptors
## Quick check on distribution of ligand-receptor weights in the network
```{r}
hist(weighted_networks_lr$weight, breaks=50)
```
```{r}
getFrequency <- function(cluster){
  cluster_cells <- colnames(seurat)[seurat$seurat_clusters == cluster]
  
  subset_mat <- as.matrix(seurat[["RNA"]]@data[, cluster_cells])
  subset_frq <- rowSums(subset_mat > 0) / ncol(subset_mat)
  genes_keep <- rownames(subset_mat)[subset_frq >= 0.1]
  return(genes_keep)
}
```


## Plot receptor weights
```{r}
getFrequency <- function(cluster){
  cluster_cells <- colnames(seurat)[Idents(seurat) == cluster]
  receptors <- lr_network %>% pull(to) %>% unique()
  receptors_check <- receptors[which(receptors %in% rownames(seurat))]
  
  subset_mat <- as.matrix(seurat[["RNA"]]@data[receptors_check, cluster_cells])
  subset_frq <- rowSums(subset_mat > 0) / ncol(subset_mat)
  genes_keep <- rownames(subset_mat)[subset_frq >= 0.1]
  return(genes_keep)
}
```

```{r}
receptor_interaction <- function(subset_name, ligand_heatmap, plot_width, plot_height){
  #Get all cell types
  subset_clusters <- subset_dict %>% filter(subset == subset_name) %>% pull(AllClusters)
  
  #Get ordered ligand list from heatmap object
  ligand_order <- ligand_heatmap$gtable$grobs[[3]]$label
  
  #Define target receptors
  receptors <- lr_network %>% pull(to) %>% unique()
  ##Iterate through clusters checking for frequency of expressing cells
  expressed_receptors <- lapply(subset_clusters, getFrequency)
  expressed_receptors <- unique(unlist(expressed_receptors))
  ##Filter network
  lr_network_top = lr_network %>% 
    filter(from %in% ligand_order & to %in% expressed_receptors) %>% 
    distinct(from,to)
  best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()
  
  #Find ligand-receptor network
  lr_network_top_df_large = weighted_networks_lr %>% 
    filter(from %in% ligand_order & to %in% best_upstream_receptors)
  ##Make a wide format
  lr_network_top_df = lr_network_top_df_large %>% spread("from","weight",fill = 0)
  lr_network_top_matrix = lr_network_top_df %>% 
    select(-to) %>% 
    as.matrix() %>% 
    magrittr::set_rownames(lr_network_top_df$to)
  
  #Cluster receptors
  dist_receptors = dist(lr_network_top_matrix, method = "binary")
  hclust_receptors = hclust(dist_receptors, method = "ward.D2")
  order_receptors = hclust_receptors$labels[hclust_receptors$order]
  
  #Test heatmap
  receptor_mat <- t(lr_network_top_matrix[order_receptors, ligand_order])
  
  ligand_receptor_heatmap <- pheatmap(receptor_mat,
         color = colorRampPalette(c("whitesmoke", "darkgreen"))(100),
         breaks = seq(0, 1, length.out=101),
         cluster_rows=F,
         cluster_cols=F,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         clustering_method="ward.D2",
         border_color = "black",
         filename=paste0("C:/.../", subset_name, "_receptor_activity.png"),
         width=plot_width, 
         height=plot_height)
  return(ligand_receptor_heatmap)
}
```

```{r}
receptor_all <- receptor_interaction("AllCells", ligand_all, 25, 10)
```


# Receptor expression in receivers
```{r}
getAvgExp <- function(cluster, gene_order){
  cells <- colnames(seurat)[seurat$seurat_clusters == cluster] #& seurat$GA == "GA_19.0"]
  avg <- rowMeans(as.matrix(seurat[["RNA"]]@data[gene_order,cells]))
  return(avg)
}
```

```{r}
receptor_expression_heatmap <- function(receptor_heatmap, ligand_heatmap,
                                      subset_name, plot_width, plot_height){
  
  #Get ligand list from heatmap
  gene_order <- receptor_heatmap$gtable$grobs[[2]]$label
  clusters <- ligand_heatmap$gtable$grobs[[2]]$label
  
  receptor_avg <- lapply(clusters, getAvgExp, gene_order=gene_order)
  receptor_avg <- do.call("cbind", receptor_avg)
  colnames(receptor_avg) <- clusters
  
  receptor_avg <- t(receptor_avg)
  
  ligand_exp_heatmap <- pheatmap(receptor_avg,
         color = colorRampPalette(c("whitesmoke", "purple"))(100),
         breaks = seq(0, 2, length.out=101),
         cluster_rows=F,
         cluster_cols=F,
         show_rownames=T,
         show_colnames=T,
         treeheight_row=0,
         treeheight_col=0,
         legend=F,
         border_color = "black",
         filename=paste0("C:/.../", subset_name, "_receptor_expression.png"),
         width=plot_width, 
         height=plot_height)
}
```

```{r}
receptor_expression_heatmap(receptor_all, ligand_all, "Allcells", 25, 3)
```
