---
title: "NicheNet"
output: html_notebook
---

```{r}
library(Seurat)
library(ggplot2)
library(viridis)
library(dplyr)
library(pheatmap)
library(nichenetr)
library(tidyverse)
library(maps)
```

#Read in the expression data of interacting cells:
```{r}
memory.limit(size=60000)
seuratObj = readRDS("C:/.../Full dataset.rds")
seuratObj@meta.data %>% head()
##         nGene nUMI orig.ident aggregate res.0.6 celltype nCount_RNA nFeature_RNA
## W380370   880 1611      LN_SS        SS       1    CD8 T       1607          876
## W380372   541  891      LN_SS        SS       0    CD4 T        885          536
## W380374   742 1229      LN_SS        SS       0    CD4 T       1223          737
## W380378   847 1546      LN_SS        SS       1    CD8 T       1537          838
## W380379   839 1606      LN_SS        SS       0    CD4 T       1603          836
## W380381   517  844      LN_SS        SS       0    CD4 T        840          513
```


#Visualize which cell populations are present
```{r}
seuratObj@meta.data$seurat_clusters %>% table() # note that the number of cells of some cell types is very low and should preferably be higher for a real application
## .
##     B CD4 T CD8 T    DC  Mono    NK  Treg 
##   382  2562  1645    18    90   131   199
DimPlot(seuratObj, reduction = "umap")
```

```{r}
seuratObj$celltype <- seuratObj$seurat_clusters
```


#Visualize the data 
```{r}
seuratObj@meta.data$GA %>% table()
## .
## LCMV   SS 
## 3886 1141
DimPlot(seuratObj, reduction = "umap", group.by = "GA")
```

#Read in NicheNet’s ligand-target prior model, ligand-receptor network and weighted integrated networks:
```{r}
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns
##                 CXCL1        CXCL2        CXCL3        CXCL5         PPBP
## A1BG     3.534343e-04 4.041324e-04 3.729920e-04 3.080640e-04 2.628388e-04
## A1BG-AS1 1.650894e-04 1.509213e-04 1.583594e-04 1.317253e-04 1.231819e-04
## A1CF     5.787175e-04 4.596295e-04 3.895907e-04 3.293275e-04 3.211944e-04
## A2M      6.027058e-04 5.996617e-04 5.164365e-04 4.517236e-04 4.590521e-04
## A2M-AS1  8.898724e-05 8.243341e-05 7.484018e-05 4.912514e-05 5.120439e-05

lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
head(lr_network)
## # A tibble: 6 x 4
##   from  to    source         database
##   <chr> <chr> <chr>          <chr>   
## 1 CXCL1 CXCR2 kegg_cytokines kegg    
## 2 CXCL2 CXCR2 kegg_cytokines kegg    
## 3 CXCL3 CXCR2 kegg_cytokines kegg    
## 4 CXCL5 CXCR2 kegg_cytokines kegg    
## 5 PPBP  CXCR2 kegg_cytokines kegg    
## 6 CXCL6 CXCR2 kegg_cytokines kegg

weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
weighted_networks_lr = weighted_networks$lr_sig %>% inner_join(lr_network %>% distinct(from,to), by = c("from","to"))

head(weighted_networks$lr_sig) # interactions and their weights in the ligand-receptor + signaling network
## # A tibble: 6 x 3
##   from  to     weight
##   <chr> <chr>   <dbl>
## 1 A1BG  ABCC6  0.422 
## 2 A1BG  ACE2   0.101 
## 3 A1BG  ADAM10 0.0970
## 4 A1BG  AGO1   0.0525
## 5 A1BG  AKT1   0.0855
## 6 A1BG  ANXA7  0.457
head(weighted_networks$gr) # interactions and their weights in the gene regulatory network
## # A tibble: 6 x 3
##   from  to     weight
##   <chr> <chr>   <dbl>
## 1 A1BG  A2M    0.0294
## 2 AAAS  GFAP   0.0290
## 3 AADAC CYP3A4 0.0422
## 4 AADAC IRF8   0.0275
## 5 AATF  ATM    0.0330
## 6 AATF  ATR    0.0355
```

#PERFORM THE NICHENET ANALYSIS:

#1. Define a “sender/niche” cell population and a “receiver/target” cell population present in your expression data and determine which genes are expressed in both populations
```{r}
Idents(seuratObj) <- "seurat_clusters"
```

```{r}
## receiver
receiver = c("6")
expressed_genes_receiver = get_expressed_genes(receiver, seuratObj, assay_oi = "RNA", pct = 0.10)

background_expressed_genes = expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
```

```{r}
## sender
sender_celltypes = c("0", "1", "2", "3", "4")

list_expressed_genes_sender = sender_celltypes %>% unique() %>% lapply(get_expressed_genes, seuratObj, assay_oi = "RNA", pct = 0.10) # lapply to get the expressed genes of every sender cell type separately here
expressed_genes_sender = list_expressed_genes_sender %>% unlist() %>% unique()
```

#2. Define a gene set of interest
```{r}
seurat_obj_receiver= subset(seuratObj, idents = receiver)
seurat_obj_receiver = SetIdent(seurat_obj_receiver, value = seurat_obj_receiver[["GA"]])

condition_oi = c("GA_19.0", "GA_18.2")
condition_reference = c("GA_14.1", "GA_14.3") 
  
DE_table_receiver = FindMarkers(object = seurat_obj_receiver, ident.1 = condition_oi, ident.2 = condition_reference, min.pct = 0.10) %>% rownames_to_column("gene")

geneset_oi = DE_table_receiver %>% filter(p_val_adj <= 0.05 & abs(avg_log2FC) >= 0.50) %>% pull(gene)
geneset_oi = geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

#3. Define a set of potential ligands
```{r}
ligands = lr_network %>% pull(from) %>% unique()
receptors = lr_network %>% pull(to) %>% unique()

expressed_ligands = intersect(ligands,expressed_genes_sender)
expressed_receptors = intersect(receptors,expressed_genes_receiver)

potential_ligands = lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% pull(from) %>% unique()
```

#4) Perform NicheNet ligand activity analysis
```{r}
ligand_activities = predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = potential_ligands)

ligand_activities = ligand_activities %>% arrange(-pearson) %>% mutate(rank = rank(desc(pearson)))
ligand_activities
## # A tibble: 44 x 5
##    test_ligand auroc  aupr pearson  rank
##    <chr>       <dbl> <dbl>   <dbl> <dbl>
##  1 Ebi3        0.662 0.238  0.219      1
##  2 Il15        0.596 0.160  0.109      2
##  3 Crlf2       0.560 0.160  0.0890     3
##  4 App         0.499 0.134  0.0750     4
##  5 Tgfb1       0.498 0.134  0.0631     5
##  6 Ptprc       0.539 0.142  0.0602     6
##  7 H2-M3       0.526 0.149  0.0533     7
##  8 Icam1       0.544 0.134  0.0496     8
##  9 Cxcl10      0.536 0.134  0.0457     9
## 10 Adam17      0.517 0.129  0.0378    10
## # ... with 34 more rows
```

#Save the ligant activities table
```{r}
write.csv(ligand_activities, file="C:/.../ligand activities table.csv", quote=F)
```

#Read the altered ligand activities table
```{r}
ligand_activities <- read.csv("C:/.../ligand activities table_processed.csv",row.names=1)
```



#The number of top-ranked ligands that are further used to predict active target genes and construct an active ligand-receptor network is here 20.
```{r}
best_upstream_ligands = ligand_activities %>% top_n(20, pearson) %>% arrange(-pearson) %>% pull(test_ligand) %>% unique()
```

#Plot the graph for quality control
```{r}
# show histogram of ligand activity scores
p_hist_lig_activity = ggplot(ligand_activities, aes(x=pearson)) + 
  geom_histogram(color="black", fill="darkorange")  + 
  # geom_density(alpha=.1, fill="orange") +
  geom_vline(aes(xintercept=min(ligand_activities %>% top_n(20, pearson) %>% pull(pearson))), color="red", linetype="dashed", size=1) + 
  labs(x="ligand activity (PCC)", y = "# ligands") +
  theme_classic()
p_hist_lig_activity
```

#These ligands are expressed by one or more of the input sender cells. To see which cell population expresses which of these top-ranked ligands, you can run the following:
```{r}
DotPlot(seuratObj, features = best_upstream_ligands %>% rev(), cols = "RdYlBu", ) + RotatedAxis()
```

#Save DotPlot
```{r}
DotPlot <- DotPlot(seuratObj, features = best_upstream_ligands %>% rev(), cols = c("light green", "dark green"), dot.scale=8) + RotatedAxis()
ggsave(DotPlot, filename="C:/.../ligand activities.png", width=10, height=15, dpi=600)
```



#5) Infer receptors and top-predicted target genes of ligands that are top-ranked in the ligand activity analysis
#Active target gene inference
```{r}
active_ligand_target_links_df = best_upstream_ligands %>% lapply(get_weighted_ligand_target_links,geneset = geneset_oi, ligand_target_matrix = ligand_target_matrix, n = 200) %>% bind_rows() %>% drop_na()

active_ligand_target_links = prepare_ligand_target_visualization(ligand_target_df = active_ligand_target_links_df, ligand_target_matrix = ligand_target_matrix, cutoff = 0.33)

order_ligands = intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev() %>% make.names()
order_targets = active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links)) %>% make.names()
rownames(active_ligand_target_links) = rownames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23
colnames(active_ligand_target_links) = colnames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23

vis_ligand_target = active_ligand_target_links[order_targets,order_ligands] %>% t()
```

#-> Visualize:
```{r}
p_ligand_target_network = vis_ligand_target %>% make_heatmap_ggplot("Prioritized ligands","Predicted target genes", color = "purple",legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential")  + theme(axis.text.x = element_text(face = "italic")) + scale_fill_gradient2(low = "whitesmoke",  high = "purple", breaks = c(0,0.0045,0.0090))
p_ligand_target_network
```

#Save Heatmap
```{r}
p_ligand_target_network = vis_ligand_target %>% make_heatmap_ggplot("Prioritized ligands","Predicted target genes", color = "purple", legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential")   + scale_fill_gradient2(low = "gray",  high = "purple", mid= "gray95", breaks = c(0,0.0025,0.005))


ggsave(p_ligand_target_network, filename="C:/.../target genes.png", width=3.5, height=5, dpi=600)
```


#Receptors of top-ranked ligands
```{r}
lr_network_top = lr_network %>% filter(from %in% best_upstream_ligands & to %in% expressed_receptors) %>% distinct(from,to)
best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()

lr_network_top_df_large = weighted_networks_lr %>% filter(from %in% best_upstream_ligands & to %in% best_upstream_receptors)

lr_network_top_df = lr_network_top_df_large %>% spread("from","weight",fill = 0)
lr_network_top_matrix = lr_network_top_df %>% select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df$to)

dist_receptors = dist(lr_network_top_matrix, method = "binary")
hclust_receptors = hclust(dist_receptors, method = "ward.D2")
order_receptors = hclust_receptors$labels[hclust_receptors$order]
    
dist_ligands = dist(lr_network_top_matrix %>% t(), method = "binary")
hclust_ligands = hclust(dist_ligands, method = "ward.D2")
order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]

order_receptors = order_receptors %>% intersect(rownames(lr_network_top_matrix))
order_ligands_receptor = order_ligands_receptor %>% intersect(colnames(lr_network_top_matrix))

vis_ligand_receptor_network = lr_network_top_matrix[order_receptors, order_ligands_receptor]
rownames(vis_ligand_receptor_network) = order_receptors %>% make.names()
colnames(vis_ligand_receptor_network) = order_ligands_receptor %>% make.names()
```

#Visualize:
```{r}
p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential")
p_ligand_receptor_network
```

#Save Heatmap
```{r}
p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_position = "top", legend_title = "Prior interaction potential")   + scale_fill_gradient2(low = "gray",  high = "mediumvioletred", mid= "gray95")
p_ligand_receptor_network

ggsave(p_ligand_receptor_network, filename="C:/.../receptors.png", width=20, height=5, dpi=600)
```


#Receptors of top-ranked ligands, but after considering only bona fide ligand-receptor interactions documented in literature and publicly available databases
```{r}
lr_network_strict = lr_network %>% filter(database != "ppi_prediction_go" & database != "ppi_prediction")
ligands_bona_fide = lr_network_strict %>% pull(from) %>% unique()
receptors_bona_fide = lr_network_strict %>% pull(to) %>% unique()

lr_network_top_df_large_strict = lr_network_top_df_large %>% distinct(from,to) %>% inner_join(lr_network_strict, by = c("from","to")) %>% distinct(from,to)
lr_network_top_df_large_strict = lr_network_top_df_large_strict %>% inner_join(lr_network_top_df_large, by = c("from","to"))

lr_network_top_df_strict = lr_network_top_df_large_strict %>% spread("from","weight",fill = 0)
lr_network_top_matrix_strict = lr_network_top_df_strict %>% select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df_strict$to)

dist_receptors = dist(lr_network_top_matrix_strict, method = "binary")
hclust_receptors = hclust(dist_receptors, method = "ward.D2")
order_receptors = hclust_receptors$labels[hclust_receptors$order]

dist_ligands = dist(lr_network_top_matrix_strict %>% t(), method = "binary")
hclust_ligands = hclust(dist_ligands, method = "ward.D2")
order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]

order_receptors = order_receptors %>% intersect(rownames(lr_network_top_matrix_strict))
order_ligands_receptor = order_ligands_receptor %>% intersect(colnames(lr_network_top_matrix_strict))

vis_ligand_receptor_network_strict = lr_network_top_matrix_strict[order_receptors, order_ligands_receptor]
rownames(vis_ligand_receptor_network_strict) = order_receptors %>% make.names()
colnames(vis_ligand_receptor_network_strict) = order_ligands_receptor %>% make.names()
```

#Visualize:
```{r}
p_ligand_receptor_network_strict = vis_ligand_receptor_network_strict %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential\n(bona fide)")
p_ligand_receptor_network_strict
```
#Save Heatmap
```{r}
p_ligand_receptor_network_strict = vis_ligand_receptor_network_strict %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential\n(bona fide)")  + scale_fill_gradient2(low = "gray",  high = "mediumvioletred", mid= "gray95")
p_ligand_receptor_network_strict


ggsave(p_ligand_receptor_network_strict, filename="C:/.../bona fide receptors.png", width=5, height=5, dpi=600)
```

#6) Add log fold change information of ligands from sender cells
```{r}
seuratObj_P14 = subset(seuratObj, GA == c("GA_19.0", "GA_18.2"))
```
```{r}
get_lfc_celltype<-function (celltype_oi, seurat_obj, condition_colname, condition_oi, 
    condition_reference, expression_pct = 0.1) 
{
    requireNamespace("Seurat")
    requireNamespace("dplyr")
    seurat_obj_celltype = SetIdent(seurat_obj, value = seurat_obj[["celltype"]])
    seuratObj_sender = subset(seurat_obj_celltype, idents = celltype_oi)
    seuratObj_sender = SetIdent(seuratObj_sender, value = seuratObj_sender[[condition_colname]])
    DE_table_sender = FindMarkers(object = seuratObj_sender, 
        ident.1 = condition_oi, ident.2 = condition_reference, 
        min.pct = expression_pct, logfc.threshold = 0.05) %>% 
        rownames_to_column("gene")
    DE_table_sender = DE_table_sender %>% as_tibble() %>% select(-p_val) %>% 
        select(gene, avg_log2FC)
    colnames(DE_table_sender) = c("gene", celltype_oi)
    return(DE_table_sender)
}
```


```{r}
# DE analysis for each sender cell type
# this uses a new nichenetr function - reinstall nichenetr if necessary!
DE_table_all = Idents(seuratObj) %>% levels() %>% intersect(sender_celltypes) %>% lapply(get_lfc_celltype, seurat_obj = seuratObj, condition_colname = "GA", condition_oi = condition_oi, condition_reference = condition_reference, expression_pct = 0.10) %>% reduce(full_join)
DE_table_all[is.na(DE_table_all)] = 0

# Combine ligand activities with DE information
ligand_activities_de = ligand_activities %>% select(test_ligand, pearson) %>% rename(ligand = test_ligand) %>% left_join(DE_table_all %>% rename(ligand = gene))
ligand_activities_de[is.na(ligand_activities_de)] = 0

# make LFC heatmap
lfc_matrix = ligand_activities_de  %>% select(-ligand, -pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities_de$ligand)
rownames(lfc_matrix) = rownames(lfc_matrix) %>% make.names()

order_ligands = order_ligands[order_ligands %in% rownames(lfc_matrix)]
vis_ligand_lfc = lfc_matrix[order_ligands,]



colnames(vis_ligand_lfc) = vis_ligand_lfc %>% colnames() %>% make.names()

p_ligand_lfc = vis_ligand_lfc %>% make_threecolor_heatmap_ggplot("Prioritized ligands","LFC in Sender", low_color = "midnightblue",mid_color = "white", mid = median(vis_ligand_lfc), high_color = "red",legend_position = "top", x_axis_position = "top", legend_title = "LFC") + theme(axis.text.y = element_text(face = "italic"))
p_ligand_lfc
```

```{r}
p_ligand_lfc = vis_ligand_lfc %>% make_threecolor_heatmap_ggplot("Prioritized ligands","LFC in Sender", low_color = "midnightblue",mid_color = "white", mid = median(vis_ligand_lfc), high_color = "red",legend_position = "top", x_axis_position = "top", legend_title = "LFC") + theme(axis.text.y = element_text(face = "italic"))
p_ligand_lfc

ggsave(p_ligand_lfc, filename="C:/.../logFC.png", width=3, height=5, dpi=600)
```

#Change colors a bit to make them more stand out
```{r}
p_ligand_lfc = p_ligand_lfc + scale_fill_gradientn(colors = c("midnightblue","blue", "grey95", "grey99","firebrick1","red"),values = c(0,0.1,0.2,0.25, 0.40, 0.7,1), limits = c(vis_ligand_lfc %>% min() - 0.1, vis_ligand_lfc %>% max() + 0.1))
p_ligand_lfc
```


#7) Summary visualizations of the NicheNet analysis
```{r}
# ligand activity heatmap
ligand_pearson_matrix = ligand_activities %>% select(pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities$test_ligand)

rownames(ligand_pearson_matrix) = rownames(ligand_pearson_matrix) %>% make.names()
colnames(ligand_pearson_matrix) = colnames(ligand_pearson_matrix) %>% make.names()

vis_ligand_pearson = ligand_pearson_matrix[order_ligands, ] %>% as.matrix(ncol = 1) %>% magrittr::set_colnames("Pearson")
p_ligand_pearson = vis_ligand_pearson %>% make_heatmap_ggplot("Prioritized ligands","Ligand activity", color = "darkorange",legend_position = "top", x_axis_position = "top", legend_title = "Pearson correlation coefficient\ntarget gene prediction ability)") + theme(legend.text = element_text(size = 9))
```
```{r}
# ligand expression Seurat dotplot
order_ligands_adapted = order_ligands
order_ligands_adapted[order_ligands_adapted == "H2.M3"] = "H2-M3" # cf required use of make.names for heatmap visualization | this is not necessary if these ligands are not in the list of prioritized ligands!
order_ligands_adapted[order_ligands_adapted == "H2.T23"] = "H2-T23" # cf required use of make.names for heatmap visualization | this is not necessary if these ligands are not in the list of prioritized ligands!
rotated_dotplot = DotPlot(seuratObj %>% subset(CellType %in% sender_celltypes), features = order_ligands_adapted %>% rev(), cols = "RdYlBu") + coord_flip() + theme(legend.text = element_text(size = 10), legend.title = element_text(size = 12)) # flip of coordinates necessary because we want to show ligands in the rows when combining all plots
```
```{r}
figures_without_legend = cowplot::plot_grid(
  p_ligand_pearson + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()),
  rotated_dotplot + theme(legend.position = "none", axis.ticks = element_blank(), axis.title.x = element_text(size = 12), axis.text.y = element_text(face = "italic", size = 9), axis.text.x = element_text(size = 9,  angle = 90,hjust = 0)) + ylab("Expression in Sender") + xlab("") + scale_y_discrete(position = "right"),
  p_ligand_lfc + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()) + ylab(""),
  p_ligand_target_network + theme(legend.position = "none", axis.ticks = element_blank()) + ylab(""),
  align = "hv",
  nrow = 1,
  rel_widths = c(ncol(vis_ligand_pearson)+6, ncol(vis_ligand_lfc) + 7, ncol(vis_ligand_lfc) + 8, ncol(vis_ligand_target)))

legends = cowplot::plot_grid(
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_pearson)),
    ggpubr::as_ggplot(ggpubr::get_legend(rotated_dotplot)),
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_lfc)),
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_target_network)),
    nrow = 1,
    align = "h", rel_widths = c(1.5, 1, 1, 1))

combined_plot = cowplot::plot_grid(figures_without_legend, legends, rel_heights = c(10,5), nrow = 2, align = "hv")
combined_plot
```


